---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# glmSparse

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of glmSparse is to provide common methods such as `print`, `summary`,
`broom::tidy`, for sparse/dense GLM models implemented in the
[MatrixModels](https://cran.r-project.org/web/packages/MatrixModels/index.html) 
package. For more information on this particular implementation of GLM models,
see the documentation for `MatrixModels::glm4()`.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("dmolitor/glmSparse")
```
## Sparse vs. Dense

The primary feature of `glmSparse()` is that it allows the user the create a sparse model
matrix and fit a generalized linear model using this sparse matrix. In cases where
the model matrix is highly sparse this will be much more memory/computationally
efficient. For a more in-depth discussion of sparse matrices and their benefits, see
[this write-up](https://cran.r-project.org/web/packages/Matrix/vignettes/sparseModels.pdf).
Using a quick example we can confirm that the results of `glm()` and `glmSparse()`
are the same.

```{r Comparison}
library(broom)
library(glmSparse)
library(Matrix)

x <- sparse.model.matrix(mpg ~ ., data = mtcars)
y <- mtcars$mpg

# stats::glm implementation
glmBase <- glm(mpg ~ ., 
               family = gaussian, 
               data = mtcars)
# glmSparse formula interface
glmSparse_form <- glmSparse(mpg ~ .,
                            family = gaussian,
                            data = mtcars)
# glmSparse alternate interface
glmSparse_alt <- glmSparse(x = x,
                           y = y,
                           family = gaussian)

# Compare the coefficient tables
tidy(glmBase)
tidy(glmSparse_form)
tidy(glmSparse_alt)
```
The results are identical!

## Why use this?

The primary reason to use this implementation is in the case of an extremely
large model matrix that is fairly sparse that will cause memory issues if stored
in standard dense format. When dealing with moderate sized datasets, there is little
to no reason to prefer this vs. the `glm()` implementation. Additionally, I have
done no comparison to the implementations found in the `speedglm` or `bigglm` packages;
both of which claim to address similar situations, potentially with better results. If
these packages indeed have faster implementations and can handle sparse model matrices
equally well, there is no reason to utilize `glmSparse()`. In other words, this is tbd.

